---
title: "WeightChange"
output: html_document
date: "2024-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(car)
library(reshape2)
```

## Load Data

```{r}
weightData <- read.table("weight_change_dataset.csv", header=TRUE, sep=",")
head(weightData)
```

## Data Cleaning

```{r}

#we treate sleep quality, physical activity level and stress level treat as continuous variables since the levels have a meaningful progression or rank (e.g., increasing physical effort or impact on health).

cleanWeight <- weightData %>%
  mutate(
    Sleep.Quality = case_when(
      Sleep.Quality == "Excellent" ~ 4,
      Sleep.Quality == "Good" ~ 3,
      Sleep.Quality == "Fair" ~ 2,
      Sleep.Quality == "Poor" ~ 1
    ),Physical.Activity.Level = case_when(
      Physical.Activity.Level == "Sedentary" ~ 1,
      Physical.Activity.Level == "Lightly Active" ~ 2,
      Physical.Activity.Level == "Moderately Active" ~ 3,
      Physical.Activity.Level == "Very Active" ~ 4
    )
  )


colnames(cleanWeight) <- c(
  "ParticipantID", "Age", "Gender", 
  "CurrentWeightLbs", "BMRCalories", "DailyCaloriesConsumed", 
  "DailyCaloricSurplusDeficit", "WeightChangeLbs", "DurationWeeks", 
  "PhysicalActivityLevel", "SleepQuality", "StressLevel", 
  "FinalWeightLbs"
)

# Convert categorical variables into factors
cleanWeight <- cleanWeight %>%
  mutate(
    Gender = factor(Gender, levels = c("M", "F"))
  )

#make ratio by weight and duration
#by gradescope suggestion 3

cleanWeight <- cleanWeight %>%
  mutate(weightChangePerWeek = WeightChangeLbs / DurationWeeks) %>%
  select(-ParticipantID, -WeightChangeLbs, -DurationWeeks)

head(cleanWeight)
```

## Check multicollinearity

```{r}
numeric_data <- cleanWeight %>%
  select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Melt the correlation matrix for visualization
cor_melt <- melt(cor_matrix)

# Create the heatmap
ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, 
                       limit = c(-1, 1), space = "Lab", name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap with Values", x = "", y = "")
```

```{r}
cleanWeight <- cleanWeight %>%
  select(-CurrentWeightLbs, -FinalWeightLbs) 

#QUESTION: should I exclude Final weight?
```

```{r}

numeric_data <- cleanWeight %>%
  select(where(is.numeric))
# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Melt the correlation matrix for visualization
cor_melt <- melt(cor_matrix)

# Create the heatmap
ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, 
                       limit = c(-1, 1), space = "Lab", name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap with Values", x = "", y = "")
```

```{r}
vif_model <- lm(weightChangePerWeek ~ ., data = cleanWeight)

# Calculate VIF
vif_values <- vif(vif_model)

# Print VIF values
print(vif_values)
```

```{r}
#Incredible High VIF mulicolinearly so delete one vairable
cleanWeight <- cleanWeight %>%
  select(-DailyCaloricSurplusDeficit)


vif_model <- lm(weightChangePerWeek ~ ., data = cleanWeight)

# Calculate VIF
vif_values <- vif(vif_model)

# Print VIF values
print(vif_values)
```

-   **VIF \< 5**: Low multicollinearity.

-   **VIF 5–10**: Moderate multicollinearity (consider investigating further).

-   **VIF \> 10**: High multicollinearity (may require remedial measures, such as removing variables or PCA).

## Start fitting, and eliminate variables

```{r}
predictors <- c("DailyCaloriesConsumed", "BMRCalories","PhysicalActivityLevel", "Age","SleepQuality", "StressLevel")

# Plot each predictor
for (var in predictors) {
  print(
    ggplot(cleanWeight, aes_string(x = var, y = "weightChangePerWeek", color = "Gender")) +
      geom_point(alpha = 0.6) +
      labs(title = paste("Interaction:", var, "vs Weight Change by Gender"),
           x = var, y = "Weight Change per Week") +
      theme_minimal()
  )
}
#It seems BMRCalories will have different trend relative to gender
```

```{r}
regWithInteraction <- lm(weightChangePerWeek ~ Age + Gender * BMRCalories + DailyCaloriesConsumed + PhysicalActivityLevel + StressLevel + SleepQuality,  cleanWeight)
regWithoutInteration <- lm(weightChangePerWeek ~ . , cleanWeight)

```

```{r}
summary(regWithoutInteraction)
#according to the p value the interaction is useless
```

```{r}

summary(regWithInteration)
```

```{r}
reg1 <- lm(weightChangePerWeek ~ Age + Gender + BMRCalories + DailyCaloriesConsumed + StressLevel + SleepQuality, cleanWeight)
summary(reg1)
```

```{r}
reg2 <- lm(weightChangePerWeek ~ Gender + BMRCalories + DailyCaloriesConsumed + StressLevel + SleepQuality, cleanWeight)
summary(reg2)
```

```{r}

reg3 <- lm(weightChangePerWeek ~  Gender + DailyCaloriesConsumed + StressLevel + SleepQuality , cleanWeight_b)
summary(reg3)
```

```{r}
RSS_model <- sum(resid(reg2)^2)         
sigma2 <- summary(reg2)$sigma^2      
p <- length(coef(reg2))                
n <- nrow(cleanWeight_a)        

Cp <- (RSS_model / sigma2) + 2 * p - n
Cp
#valid CP 5+1
plot(reg2)
```

```{r}

RSS_model <- sum(resid(reg3)^2)         
sigma2 <- summary(reg3)$sigma^2      
p <- length(coef(reg3))                
n <- nrow(cleanWeight_b)        

Cp <- (RSS_model / sigma2) + 2 * p - n
Cp
#Valid CP 4+1
plot(reg3)
```

### Previous residual plot seems shows some pattern

```{r}

meanDCC = mean(cleanWeight$DailyCaloriesConsumed)
reg_cubic <- lm(weightChangePerWeek ~ Gender + 
                 I((meanDCC-DailyCaloriesConsumed)^3) + SleepQuality + StressLevel,  data = cleanWeight)

summary(reg_cubic)
```

```{r}
plot(reg_cubic)
```

## Validation

```{r}
squared_errors <- numeric(n)

# Perform LOOCV
for (i in 1:n) {
  # Split the data: exclude the i-th observation
  train_data <- cleanWeight[-i, ]
  test_data <- cleanWeight[i, , drop = FALSE]
  
  # Fit the model on the training set
  loocv_model <- lm(weightChangePerWeek ~ Gender + 
                      I((meanDCC - DailyCaloriesConsumed)^3) + SleepQuality + StressLevel,
                    data = train_data)
  
  # Predict the response for the left-out observation
  predicted <- predict(loocv_model, newdata = test_data)
  
  # Calculate the squared error for the i-th observation
  squared_errors[i] <- (test_data$weightChangePerWeek - predicted)^2
}

# Compute the LOOCV MSE
loocv_mse <- mean(squared_errors)
cat("LOOCV Mean Squared Error:", loocv_mse, "\n")
```

LOOCV is a test that treat 1 data as the test set and 99 as the training set,

since we have 100 data, so each data has a chance to become the "test set"
