---
title: "WeightChange"
output: html_document
date: "2024-11-24"
---

```{r}
library(leaps)
library(dplyr)
library(ggplot2)
library(car)
library(reshape2)
```

## Load Data

```{r}
weightData <- read.table("weight_change_dataset.csv", header=TRUE, sep=",")
head(weightData)
```

## Data Cleaning

```{r}

#we treate sleep quality and stress level treat as continuous variables since the levels have a meaningful progression or rank (e.g., increasing physical effort or impact on health).

cleanWeight <- weightData %>%
  mutate(
    Sleep.Quality = case_when(
      Sleep.Quality == "Excellent" ~ 4,
      Sleep.Quality == "Good" ~ 3,
      Sleep.Quality == "Fair" ~ 2,
      Sleep.Quality == "Poor" ~ 1
    )
  )


colnames(cleanWeight) <- c(
  "ParticipantID", "Age", "Gender", 
  "CurrentWeightLbs", "BMRCalories", "DailyCaloriesConsumed", 
  "DailyCaloricSurplusDeficit", "WeightChangeLbs", "DurationWeeks", 
  "PhysicalActivityLevel", "SleepQuality", "StressLevel", 
  "FinalWeightLbs"
)

# Convert categorical variables into factors
cleanWeight <- cleanWeight %>%
  mutate(
    Gender = factor(Gender, levels = c("M", "F"))
  )

#make ratio by weight and duration
#by gradescope suggestion 3

cleanWeight <- cleanWeight %>%
  mutate(weightChangePerWeek = WeightChangeLbs / DurationWeeks) %>%
  select(-ParticipantID, -WeightChangeLbs, -DurationWeeks)

head(cleanWeight)
```

# See need transform or not

```{r}
model <- lm(weightChangePerWeek ~ ., data = cleanWeight)
plot(model$fitted.values, residuals(model), 
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", 
     ylab = "Residuals")
abline(h = 0, col = "red")
```

```{r}
hist(cleanWeight$weightChangePerWeek, 
     main = "Histogram of weightChangePerWeek", 
     xlab = "Weight Change Per Week")
```

obvious pattern, left skewed need transform

```{r}
model <- lm(exp(weightChangePerWeek) ~ ., data = cleanWeight)
plot(model$fitted.values, residuals(model), 
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", 
     ylab = "Residuals")
abline(h = 0, col = "red")
```

```{r}
expWeight <- cleanWeight %>%
  mutate(expWeightCPW = exp(weightChangePerWeek)) %>%
  subset(select = -c(weightChangePerWeek) )
```

## Check multicollinearity

```{r}
numeric_data <- expWeight %>%
  select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Melt the correlation matrix for visualization
cor_melt <- melt(cor_matrix)

# Create the heatmap
ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, 
                       limit = c(-1, 1), space = "Lab", name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap with Values", x = "", y = "")
```

```{r}
vif_model <- lm(expWeightCPW ~ ., data = expWeight)

# Calculate VIF
vif_values <- vif(vif_model)

# Print VIF values
print(vif_values)
```

```{r}
#Incredible High VIF mulicolinearly so delete one vairable
expWeight <- expWeight %>%
  select(-DailyCaloriesConsumed)


vif_model <- lm(expWeightCPW ~ ., data = expWeight)

# Calculate VIF
vif_values <- vif(vif_model)

# Print VIF values
print(vif_values)
```

```{r}
expWeight <- expWeight %>%
  select(-BMRCalories)


vif_model <- lm(expWeightCPW ~ ., data = expWeight)

# Calculate VIF
vif_values <- vif(vif_model)

# Print VIF values
print(vif_values)
```

```{r}
expWeight <- expWeight %>%
  select(-FinalWeightLbs)


vif_model <- lm(expWeightCPW ~ ., data = expWeight)

# Calculate VIF
vif_values <- vif(vif_model)

# Print VIF values
print(vif_values)
```

-   **VIF \< 5**: Low multicollinearity.

-   **VIF 5–10**: Moderate multicollinearity (consider investigating further).

-   **VIF \> 10**: High multicollinearity (may require remedial measures, such as removing variables or PCA).

# Start Fiting

```{r}
regAll <- lm(expWeightCPW ~ .,  expWeight)
summary(regAll)
```

```{r}
regAll %>%
  ggplot(aes(x = regAll$fitted.values , y = regAll$residuals)) +
  geom_point() +
   labs(title = "Residuals vs. Fitted value",
       x = "fitted values", y = "residuals")
```

```{r}
ggplot(expWeight, aes(x = as.factor(PhysicalActivityLevel), y = expWeightCPW, fill = as.factor(PhysicalActivityLevel))) +
  geom_boxplot() +
  labs(title = "Boxplot of Weight Change Per Week by Physical Activity Level",
       x = "Physical Activity Level",
       y = "Weight Change Per Week") +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal()
```

With boxplot we can see that the median difference between levels is not large. Lightly Active and Sedentary median close together, and Sedentary p value high as 0.9, combine with baseline.

```{r}
newCleanWeight <- expWeight

newCleanWeight$PhysicalActivityLevel <- as.character(newCleanWeight$PhysicalActivityLevel) 
newCleanWeight$PhysicalActivityLevel[newCleanWeight$PhysicalActivityLevel %in% c("Sedentary", "Lightly Active")] <- "low Activity"
newCleanWeight$PhysicalActivityLevel[newCleanWeight$PhysicalActivityLevel %in% c("Moderately Active", "Very Active")] <- "high Activity"

newCleanWeight$PhysicalActivityLevel <- as.factor(newCleanWeight$PhysicalActivityLevel)

regAll <- lm(expWeightCPW ~ .,  newCleanWeight)
summary(regAll)
newCleanWeight
```

```{r}

# s = regsubsets(expWeightCPW~., data=newCleanWeight, method="exhaustive")
# 
# ss <- summary(s)
# ss$outmat
```

```{r}
# num_predictors <- rownames(ss$which)
# 
# 
# plot(num_predictors, ss$rsq, type = "b", pch = 19,
#      main = "R^2 vs Number of Predictors",
#      xlab = "Number of Predictors", ylab = "R^2")
# 
# # Plot Adjusted R^2
# plot(num_predictors, ss$adjr2, type = "b", pch = 19,
#      main = "Adjusted R^2 vs Number of Predictors",
#      xlab = "Number of Predictors", ylab = "Adjusted R^2")
# 
# 
# num_predictors <- 1:length(ss$cp) 
# cp_values <- ss$cp
# 
# # Plot Cp 
# plot(num_predictors, cp_values, type = "b", pch = 19,
#      main = "Mallows' Cp vs Number of Predictors",
#      xlab = "Number of Predictors", ylab = "Cp")
# 
# abline(a = 1, b = 1, col = "royalblue", lty = 2, lwd = 2)
# 
# text(num_predictors, cp_values, labels = round(cp_values, 2), pos = 3, cex = 0.8, col = "black")

```

# Choose model

test based, eliminate variable through p value

```{r}
regfull <- lm(expWeightCPW ~ DailyCaloricSurplusDeficit + SleepQuality +  StressLevel + Age + Gender + CurrentWeightLbs + PhysicalActivityLevel,  newCleanWeight)
summary(regfull)
```

```{r}

regfull %>%
  ggplot(aes(x = regfull$fitted.values , y = regfull$residuals)) +
  geom_point() +
   labs(title = "Residuals vs. Fitted value",
       x = "fitted values", y = "residuals")
```

```{r}
reg6 <- lm(expWeightCPW ~ DailyCaloricSurplusDeficit + SleepQuality +  StressLevel + Age + Gender + CurrentWeightLbs,  newCleanWeight)
summary(reg6)
```

```{r}
reg6 %>%
  ggplot(aes(x = reg6$fitted.values , y = reg6$residuals)) +
  geom_point() +
   labs(title = "Residuals vs. Fitted value",
       x = "fitted values", y = "residuals")
```

```{r}
reg5<- lm(expWeightCPW ~ DailyCaloricSurplusDeficit + SleepQuality +  StressLevel + Age + Gender,  newCleanWeight)
summary(reg5)
```

```{r}
reg5 %>%
  ggplot(aes(x = reg5$fitted.values , y = reg5$residuals)) +
  geom_point() +
   labs(title = "Residuals vs. Fitted value",
       x = "fitted values", y = "residuals")
```

```{r}
reg4<- lm(expWeightCPW ~ DailyCaloricSurplusDeficit + SleepQuality +  StressLevel + Gender,  newCleanWeight)
summary(reg4)
```

```{r}
reg4 %>%
  ggplot(aes(x = reg6$fitted.values , y = reg4$residuals)) +
  geom_point() +
   labs(title = "Residuals vs. Fitted value",
       x = "fitted values", y = "residuals")
```

```{r}
reg3<- lm(expWeightCPW ~ DailyCaloricSurplusDeficit + SleepQuality +  StressLevel,  newCleanWeight)
summary(reg3)
```

```{r}
ggplot(reg3, aes(x = reg3$fitted.values, y = reg3$residuals)) +
  geom_point() +
  labs(title = "Residuals vs. Fitted value",
       x = "Fitted Value", y = "Residuals") +
  theme_minimal()

```

### 6. **Check for Interaction Terms**

-   Consider adding interaction terms (`x1 * x2`) if theory or plots suggest interactions.

-   Test interaction significance using p-values.

### 7. **Check for Polynomial or Quadratic Terms**

-   If scatterplots or residuals suggest curvature, consider adding quadratic terms (`x^2`).

-   Fit models with polynomial terms and compare using AIC, BIC, or adjusted R2R\^2R2.

### 8. **Model Diagnostics and Final Assessment**

-   **Residual Analysis**: Ensure residuals are normally distributed, have constant variance, and are independent.

-   **Outliers and Influential Points**: Check Cook’s distance or leverage points.

-   **Compare Models**: Use cross-validation or validation datasets.

We look at the scatterplots to see if there might be an interaction between the variables.

```{r}
ggplot(newCleanWeight, aes(x = StressLevel, y = expWeightCPW, color = as.factor(PhysicalActivityLevel))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "PhysicalActivityLevel and SleepQuality",
       x = "Stress Level",
       y = "Weight Change Per Week")
```

```{r, message=FALSE}
ggplot(newCleanWeight, aes(x = SleepQuality, y = expWeightCPW, color = as.factor(PhysicalActivityLevel))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "PhysicalActivityLevel and SleepQuality",
       x = "Sleep Quality",
       y = "Weight Change Per Week")
```

The slope of the fitted line does not change significantly, and we can test the significance of the interaction term to determine whether to include interaction.

```{r}
regLab <- lm(exp(expWeightCPW) ~ PhysicalActivityLevel * StressLevel + SleepQuality ,  newCleanWeight)
summary(regLab)
```

```{r}
regLab <- lm(exp(expWeightCPW) ~ PhysicalActivityLevel * SleepQuality +  StressLevel,  newCleanWeight)
summary(regLab)
```

All of the interaction terms are not significant, so we don't include the interaction.

```{r}
ggplot(data = cleanWeight, aes(x = SleepQuality, y = weightChangePerWeek)) +
  geom_point() +
  labs(title = "Linear vs Quadratic Fit", x = "Sleep Quality", y = "weight Change Per Week")

ggplot(data = cleanWeight, aes(x = StressLevel, y = weightChangePerWeek)) +
  geom_point() +
  labs(title = "Linear vs Quadratic Fit", x = "Stress Level", y = "weight Change Per Week")
```

The scatterplot shows that Sleep Quality, Stress Level and weight Change Per Week do not have an obvious quadratic shape, so we still choose the linear model.

## Validation

cross validation, r squre, adjust r square, mallow cp...

```{r}
n = nrow(cleanWeight)
squared_errors <- numeric(n)

# Perform LOOCV
for (i in 1:n) {
  train_data <- cleanWeight[-i, ]
  test_data <- cleanWeight[i, , drop = FALSE]
  predicted <- predict(reg4, newdata = test_data)
  squared_errors[i] <- (test_data$weightChangePerWeek - log(predicted))^2
}

loocv_mse <- mean(squared_errors)
cat("LOOCV Mean Squared Error:", loocv_mse, "\n")
```

LOOCV is a test that treat 1 data as the test set and 99 as the training set,

since we have 100 data, so each data has a chance to become the "test set"
